Haskell in Computational Biology

### 2013-05-17 jfp.tex
work on draft

A~model is represented as a sequence of \emph{nodes};
node~$j$ includes states $M_j$, $I_j$, and~$D_j$,
as well as the probabilities of transitions out of that node.
Each node contains tables of emission scores
$e'_{M_j}$~and~$e'_{I_j}$.
These tables are read by function
\texttt{eScore}, whose specification
is % parallel structure for line breaking below
\mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$}.
%
We place the transition probabilities 
into a record in which each field is labeled
\texttt{$s$\_$\mskip 1mu\hat s$},
where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
labels: 
\smallverbatiminput{tprob-tprobs}
The fields of this record are read by
function \texttt{aScore}, 
which is given states $s$~and~$\hat s$ and a node index; its %% fixes a bad line break
specification is
\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$}.

@@

@@

@@

\subsection{Software deltas}

Original Viterbi took a state and two indices as argument, and it
dealt with impermissible transitions by assigning a very large score
(which translates to a very small probability).
Indices 1, 0, and $-1$ were all used to indicate special conditions,
and 
the function had 25~cases overall.

>>>>

A~model is represented as a sequence of \emph{nodes};
node~$j$ includes states $M_j$, $I_j$, and~$D_j$,
as well as the probabilities of transitions out of that node.
Each node contains tables of emission scores
$e'_{M_j}$~and~$e'_{I_j}$.
These tables are read by function
\texttt{eScore}, whose specification
is % parallel structure for line breaking below
\mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$}.
%
We place the transition probabilities 
into a record in which each field is labeled
\texttt{$s$\_$\mskip 1mu\hat s$},
where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
labels:\label{sec:tprobs}
\smallverbatiminput{tprob-tprobs}
\remark{Need to explain those last two fields. HMMer?}
The fields of this record are read by
function \texttt{aScore}, 
which is given states $s$~and~$\hat s$ and a node index; its %% fixes a bad line break
specification is
\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$}.

@@

\subsection{Competing with C++; less FORTRAN; more Haskell}

EXPLAIN SECTION TITLE.

WHY IS THIS SECTION RELEVANT?  TO SHOW TRAPS AND PITFALLS FOR NEW
HASKELL PROGRAMMERS.

FIRST STEP WAS TO SIMPLIFY THE CODE.

As shown in \figrefpair{viterbi-original}{viterbi-transformed} and in
the code for \verb+vee'+ on page~\pageref{code:vee-prime}, Viterbi's
$V'$ is indexed by a residue index~$i$, a~node index~$j$,
and
a state $M$, $I$, or~$D$.
A~query sequence is simply an array of~\texttt{AA} indexed by~$i$,
and a model is an array of nodes indexed by~$j$.
The node array and its indices are handled as in our group's C++ code:
all nodes have the same representation, and every node can be
interpreted in one of the special states Begin and End.
(This possible interpretation causes the introduction of the
mysterious probabilities \verb+b_m+ and \verb+m_e+ into the transition
table on page~\pageref{sec:tprobs}.)
These conventions of representation lead to a full implementation of
Viterbi's algorithm that is more reminiscent of FORTRAN than of
Haskell:
\begin{itemize}
\item
Recursive computations are done on smaller array indices, not smaller
models.
\item
Array indices 1, 0, and $-1$ are all used to indicate special
conditions.
\item
Senseless combinations of array indices and states are handled by
assigning each such combination the largest representable score, which
corresponds to the smallest representable probability.
\end{itemize}
This is the code in which \mrfy\ spends 98\%\ of its time, and that
we~wanted to speed up to make it more competitive with~C++.
But the central function \verb+vee'+ has 25~cases overall.
Before thinking seriously about how to tune it, we~decided to simplify~it.

Our first step in simplification was to start thinking like Haskell
programmers and make more effective use of both the static type system:
(AND THE PATTERN-MATCH COMPILER)
\begin{itemize}
\item
We introduced a \texttt{Model} type that consists of a single Begin
node, an~unbounded sequence of Middle nodes, and a single End node.
\item
We~observed that the main algorithm where \mrfy\ spends all its time
uses only Match, Insert, and Delete states.
We~redefined the \texttt{StateLabel} type to include only these
states; the Begin and End labels are now part of
type \texttt{FullStateLabel}. 
\end{itemize}
As a result of these changes, in the revised code,
\begin{itemize}
\item
Only array index~0 is used to indicate a special
conditions: the transition from the leftmost middle node to the unique
begin node.
\item
All well-typed combinations of array indices and states make sense;
senseless combinations are ruled out by the static type system.
\end{itemize}

\begin{figure}
\def\vsum#1#2{#1+#2}
\[
V_{j}^{\prime \hat s}(i) = e^{\prime}_{\hat s_{j}}(x_{i}) 
  + \min\limits_s \left\{
       \vsum{V_{j-1}^{\prime s}(i - 1)} {a^{\prime}_{s \,\hat
       s}}, \mbox{ where $s$ can precede $\hat s$} \right \}
\]

\caption{Viterbi's transformed equations, abstracted over states $s$
and $\hat s$, defining $e^{\prime}_D(x) = 0$}
\figlabel{viterbi-abstracted}
\end{figure}

We also

The right-hand sideOur 
Indices 1, 0, and $-1$ were all used to indicate special conditions,
and

@@

NEW PARAGRAPH: STEP SLOWLY AWAY FROM YOUR CODE, AND SEE IF YOU THINK
IT RESEMBLES ANYTHING A HASKELL PROGRAMMER WOULD HAVE WRITTEN.

@@
